New cardinality estimation algorithms for HyperLogLog sketches

### 2017-01-01 paper/paper.tex
changed format of special function names

\subsection{Maximum likelihood estimation algorithm}
In order to get a fast cardinality estimation algorithm, it is crucial to minimize evaluation costs for \eqref{equ:funcprime}. A couple of optimizations allow significant reduction of computational effort:
\begin{itemize}
\item Only a fraction of all count values $\symCountVariate_\symRegVal$ is non-zero. If we denote $\symRegValVariate_\text{min}:=\min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$ and $\symRegValVariate_\text{max}:=\max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegValVariate_\text{min}, \symRegValVariate_\text{max}]$.
\item The sum $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}$ in \eqref{equ:funcprime} can be precalculated and reused for all function evaluations.
\item Many programming languages allow the efficient multiplication and division by any integral power of two using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}},\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$. This series corresponds to a geometric series with ratio two. A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function. However, if we know $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
\begin{equation}
\label{equ:helper_recursion1}
\symHelper(4\symX) = \frac{\symX+\symHelper(2\symX)\left(1-\symHelper(2\symX)\right)}{\symX+\left(1-\symHelper(2\symX)\right)}.
\end{equation}
%or
%\begin{equation}
%\label{equ:helper_recursion2}
%\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
%\end{equation}
This recursive formula is stable in a sense that the relative error of $\symHelper(4\symX)$ is smaller than that of $\symHelper(2\symX)$ as shown in \cref{app:helper_stable}.

@@

The calculation of \eqref{equ:max_likelihood_phi_func} and its derivatives requires some care when calculating $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$. Since $\symX_{\ast\symRegVal}$ is nonnegative, we have $\symY_{\ast\symRegVal},\symZ_{\ast\symRegVal}\in[0,1]$. In order to reduce the numerical error of $\symZ_{\ast\symRegVal}$ for small $\symX_{\ast\symRegVal}$, it is essential to use the function $\texttt{expm1}(\symX):=e^\symX-1$ that is available in most programming languages. If $\symX_{\ast\symRegVal}$ is smaller than $\log(2)$, we calculate $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$ via $\symZ_{\ast\symRegVal} = 
-\texttt{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$ and not as defined in \eqref{equ:definition_xyz}. In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and still only a single exponential function needs to be evaluated.

>>>>

\subsection{Maximum likelihood estimation algorithm}
In order to get a fast cardinality estimation algorithm, it is crucial to minimize evaluation costs for \eqref{equ:funcprime}. A couple of optimizations allow significant reduction of computational effort:
\begin{itemize}
\item Only a fraction of all count values $\symCountVariate_\symRegVal$ is non-zero. If we denote $\symRegValVariate_\text{min}:=\min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$ and $\symRegValVariate_\text{max}:=\max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegValVariate_\text{min}, \symRegValVariate_\text{max}]$.
\item The sum $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}$ in \eqref{equ:funcprime} can be precalculated and reused for all function evaluations.
\item Many programming languages allow the efficient multiplication and division by any integral power of two using special functions, such as $\operatorname{ldexp}$ in C/C++ or $\operatorname{scalb}$ in Java.
\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}},\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$. This series corresponds to a geometric series with ratio two. A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function. However, if we know $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
\begin{equation}
\label{equ:helper_recursion1}
\symHelper(4\symX) = \frac{\symX+\symHelper(2\symX)\left(1-\symHelper(2\symX)\right)}{\symX+\left(1-\symHelper(2\symX)\right)}.
\end{equation}
%or
%\begin{equation}
%\label{equ:helper_recursion2}
%\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
%\end{equation}
This recursive formula is stable in a sense that the relative error of $\symHelper(4\symX)$ is smaller than that of $\symHelper(2\symX)$ as shown in \cref{app:helper_stable}.

@@

The calculation of \eqref{equ:max_likelihood_phi_func} and its derivatives requires some care when calculating $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$. Since $\symX_{\ast\symRegVal}$ is nonnegative, we have $\symY_{\ast\symRegVal},\symZ_{\ast\symRegVal}\in[0,1]$. In order to reduce the numerical error of $\symZ_{\ast\symRegVal}$ for small $\symX_{\ast\symRegVal}$, it is essential to use the function $\operatorname{expm1}(\symX):=e^\symX-1$ that is available in most programming languages. If $\symX_{\ast\symRegVal}$ is smaller than $\log(2)$, we calculate $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$ via $\symZ_{\ast\symRegVal} = 
-\operatorname{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$ and not as defined in \eqref{equ:definition_xyz}. In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and still only a single exponential function needs to be evaluated.
